---
title: "Interactive ROI Histograms"
output:
  html_document:
    self_contained: true
    theme: lumen
    toc: true
    toc_depth: 2
    toc_float: true
params:
  site: "My Cohort"
  rois_list: null
  data_table: null
  hbins: 30
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plotly)
library(htmlwidgets)

# Retrieve parameters
site       <- params$site
rois_list  <- params$rois_list$value
data_table <- params$data_table
hbins      <- params$hbins

# --- Sanity Checks ---
if (is.null(data_table) || !is.data.frame(data_table)) {
  stop("Error: Data table not provided or is not a data frame in the R Markdown template.")
}

# Ensure only valid ROI columns are used
rois_list <- intersect(rois_list, names(data_table))
```

---

# Histograms
This report provides **interactive histograms** for each region of interest (ROI) in your dataset.  
Use your mouse to **zoom**, **pan**, and **hover** over green (< 2sd) or red (≥  5sd) bars to identify subject ids of outliers. (These should be visually inspected for potential issues)

**Study:** `r site`  
**Number of ROIs processed:** `r length(rois_list)`

---

```{r html_inputs, results='asis'}

html_content <- '
<div id="graphSettings" 
     style="background-color: #f8f9fa; border: 1px solid #ddd; border-radius: 10px; 
            padding: 12px 16px; margin: 12px 0 20px 0; 
            font-family: Arial, sans-serif; box-shadow: 0 2px 6px rgba(0,0,0,0.05);">

  <h3 style="margin: 0 0 10px 0; font-size: 18px; color: #333; border-bottom: 1px solid #e0e0e0; padding-bottom: 4px;">
    Graph Settings
  </h3>
  
  <div style="display:flex; align-items: center; justify-content: space-around; gap: 2rem">
  
  <div>
    <label for="scaleInput" style="font-size: 14px; color: #444; font-weight: bold;">Y-axis Scale:</label>
    <select id="scaleInput" 
            style="padding: 5px 10px; border: 1px solid #ccc; border-radius: 6px; 
                   background-color: #fff; font-size: 14px; cursor: pointer; 
                   transition: all 0.2s;">
      <option value="linear" selected>Linear</option>
      <option value="log">Log</option>
    </select>
  </div>
  
  <div>
    <label for="sdLines" style="font-size: 14px; color: #444; font-weight: bold;">Toggle Lines:</label>
    <input type="checkbox" id="sdLines" checked>
  </div>
  <div>
  
    <label for="legend" style="font-size: 14px; color: #444; font-weight: bold;">Toggle Legend:</label>
    <input type="checkbox" id="legend" checked>
  </div>
  
  </div>
</div>

'
htmltools::HTML(html_content)
```

```{r helpers, include=FALSE}
# --- Compute Basic Statistics ---
compute_roi_stats <- function(DATA) {
  list(
    N     = sum(!is.na(DATA)),
    mean  = mean(DATA, na.rm = TRUE),
    sd    = sd(DATA, na.rm = TRUE),
    min   = min(DATA, na.rm = TRUE),
    max   = max(DATA, na.rm = TRUE)
  )
}

# --- Prepare Histogram Data ---
prepare_histogram_data <- function(DATA, hbins, mu, sdev, sd_multiples = c(2, 5)) {
  hist_obj   <- hist(DATA, breaks = hbins, plot = FALSE)
  bin_breaks <- hist_obj$breaks
  bin_counts <- hist_obj$counts
  bin_mids   <- (bin_breaks[-length(bin_breaks)] + bin_breaks[-1]) / 2
  bin_widths <- diff(bin_breaks)

  plot_df <- data.frame(
    bin_mids   = bin_mids,
    bin_counts = bin_counts,
    bin_widths = bin_widths,
    bin_start  = bin_breaks[-length(bin_breaks)],
    bin_end    = bin_breaks[-1]
  )

  sorted_sd_multiples_asc <- sort(as.numeric(sd_multiples), decreasing = FALSE)
  min_sd_multiple <- sorted_sd_multiples_asc[1]
  max_sd_multiple <- sorted_sd_multiples_asc[length(sorted_sd_multiples_asc)]

  plot_df$type <- NA_character_

  lower_bound_min <- mu - min_sd_multiple * sdev
  upper_bound_min <- mu + min_sd_multiple * sdev
  plot_df$type[plot_df$bin_mids >= lower_bound_min & plot_df$bin_mids <= upper_bound_min] <- paste0("< ", min_sd_multiple, "σ")

  if (length(sorted_sd_multiples_asc) > 1) {
    for (i in 1:(length(sorted_sd_multiples_asc) - 1)) {
      current_m <- sorted_sd_multiples_asc[i]     # e.g., 2
      next_m    <- sorted_sd_multiples_asc[i + 1] # e.g., 5

      lower_current <- mu - current_m * sdev
      upper_current <- mu + current_m * sdev
      lower_next    <- mu - next_m * sdev
      upper_next    <- mu + next_m * sdev

      plot_df$type[plot_df$bin_mids >= lower_next & plot_df$bin_mids < lower_current] <- paste0("≥ ", current_m, "σ")

      plot_df$type[plot_df$bin_mids > upper_current & plot_df$bin_mids <= upper_next] <- paste0("≥ ", current_m, "σ")
    }
  }
  
   # --- adjust bins that straddle σ boundaries upward ---
  for (m in sorted_sd_multiples_asc) {
    lower_bound <- mu - m * sdev
    upper_bound <- mu + m * sdev

    crosses_lower <- plot_df$bin_start < lower_bound & plot_df$bin_end > lower_bound
    crosses_upper <- plot_df$bin_start < upper_bound & plot_df$bin_end > upper_bound

    plot_df$type[crosses_lower | crosses_upper] <- paste0("≥ ", m, "σ")
  }
  

  plot_df$type[is.na(plot_df$type)] <- paste0("≥ ", max_sd_multiple, "σ")

  plot_df
}

# --- Add Hover Text for Bins ---
add_hover_text <- function(plot_df, DATA, SUBJECT_DATA = NULL, mu, sedv, sd_multiples = c(2, 5)) {
  plot_df$hover_text <- ""
  SUBJECT_IDS_AVAILABLE <- !is.null(SUBJECT_DATA)

  min_sd_multiple <- min(sd_multiples)
  innermost_type_name <- paste0("< ", min_sd_multiple, "σ")
  sd_boundaries <-  c(mu + sd_multiples * sedv, mu - sd_multiples * sedv)

  # Ensure sd_multiples is sorted for correct "next_sd_val" determination
  sd_multiples <- sort(unique(sd_multiples))

  for (i in seq_len(nrow(plot_df))) {
    bin_min <- plot_df$bin_start[i]
    bin_max <- plot_df$bin_end[i]

    if (plot_df$type[i] != innermost_type_name && SUBJECT_IDS_AVAILABLE) {

      # Base condition
      in_bin <- DATA >= bin_min & DATA < bin_max

      # Is Bin straddling sd_line
      crossing_sd <- sd_boundaries[sd_boundaries > bin_min & sd_boundaries < bin_max]

      if(length(crossing_sd) > 0) {
        boundary <- crossing_sd[which.min(abs(crossing_sd - mu))]

        # Check if +ve or -ve boundary
        if(boundary > mu){
          in_bin <- in_bin & DATA >= boundary
        } else {
          in_bin <- in_bin & DATA <= boundary
        }
      }

      subjects_in_bin <- unique(SUBJECT_DATA[in_bin])

      if (length(subjects_in_bin) > 0) {
        if(length(subjects_in_bin) >= 100) {
          hover_txt <- paste0("Frequency: ", plot_df$bin_counts[i], "<br> <i>Note: No. of subjects in this bin are too large and cannot be displayed here.</i>")
        } else {
          current_type_string <- plot_df$type[i] # e.g., "≥ 2σ" or "≥ 5σ"
          new_label_for_hover <- ""

          sd_val_match <- regmatches(current_type_string, regexec("≥\\s*([0-9.]+)\\s*σ", current_type_string))

          if (length(sd_val_match[[1]]) > 1) {
            current_sd_val <- as.numeric(sd_val_match[[1]][2])

            current_sd_index <- which(sd_multiples == current_sd_val)

            if (length(current_sd_index) > 0) {
              current_sd_index <- current_sd_index[1]

              if (current_sd_index == length(sd_multiples)) {
                # Last SD multiple 
                new_label_for_hover <- paste0("Subjects: ", current_type_string)
              } else {
                # Intermediate SD multiple
                next_sd_val <- sd_multiples[current_sd_index + 1]
                new_label_for_hover <- paste0("Subjects: ", current_type_string, " and < ", next_sd_val, "σ")
              }
            } else {
              new_label_for_hover <- paste0("Subjects: ", current_type_string)
            }
          } else {
             new_label_for_hover <- paste0("Subjects: ", current_type_string)
          }

          hover_txt <- paste0("<b>", new_label_for_hover, "</b><br>", paste0(subjects_in_bin, collapse = "\n"))
        }
      } else {
        hover_txt <- paste0("Frequency: ", plot_df$bin_counts[i], "<br>No subjects identified in this bin.")
      }
    } else {
      hover_txt <- paste0("Frequency: ", plot_df$bin_counts[i])
    }
    plot_df$hover_text[i] <- hover_txt
  }

  plot_df
}
# --- Generate Interactive Plot ---
generate_roi_plot <- function(plot_df, roi_name, site, bin_breaks, mu, sdev, y_axis_type_param = 'linear', sd_multiples = c(2, 5)) {
  

  sdev_lines_multiples <- c(-rev(sd_multiples), sd_multiples) # e.g., for c(2,5) -> c(-5, -2, 2, 5)
  sdev_lines_x <- mu + sdev_lines_multiples * sdev
  
  sdev_shapes <- list()
  sdev_annotations <- list()

  for (i in seq_along(sdev_lines_multiples)) {
    x_val <- sdev_lines_x[i]
    multiple <- sdev_lines_multiples[i]
    
    if (x_val >= min(bin_breaks) && x_val <= max(bin_breaks)) {
      sdev_shapes[[length(sdev_shapes) + 1L]] <- list(
        type = "line",
        x0 = x_val,
        x1 = x_val,
        y0 = 0,
        y1 = 1,
        yref = "paper",
        line = list(color = "maroon", dash = "dot", width = 1)
      )

      sdev_annotations[[length(sdev_annotations) + 1L]] <- list(
        x = x_val,
        y = 0.98,
        xref = "x",
        yref = "paper",
        text = paste0(ifelse(multiple >=  0, paste0("+", multiple, "σ"), paste0(multiple, "σ"))),
        showarrow = FALSE,
        xanchor = "center",
        yanchor = "top", 
        font = list(color = "maroon", size = 16)
      )
    }
  }
  
  sdev_shapes <- Filter(Negate(is.null), sdev_shapes) 
  
  all_bin_types <- unique(plot_df$type)
  
  plotly_colors <- list()

  min_sd_mult_str <- paste0("< ", min(sd_multiples), "σ")
  max_sd_mult_str <- paste0("≥ ", max(sd_multiples), "σ")

  if (min_sd_mult_str %in% all_bin_types) {
    plotly_colors[[min_sd_mult_str]] <- "#749F9A"
  }
  if (max_sd_mult_str %in% all_bin_types) {
    plotly_colors[[max_sd_mult_str]] <- "#936269"
  }

  intermediate_bands <- setdiff(all_bin_types, c(min_sd_mult_str, max_sd_mult_str))
  
  if (length(intermediate_bands) > 0) {
    sorted_intermediate_bands <- sort(intermediate_bands) 
    
    available_band_colors <- c("#6D83A3", "#a9ddd6", "#91adc2", "#7a8b99", "#9ba0bc", "#c1b8c8")
    
    for (idx in seq_along(sorted_intermediate_bands)) {
      band_type <- sorted_intermediate_bands[idx]
      plotly_colors[[band_type]] <- available_band_colors[((idx - 1) %% length(available_band_colors)) + 1]
    }
  }
  
  plotly_colors_vector <- unlist(plotly_colors)
  names(plotly_colors_vector) <- names(plotly_colors)
  
  plot_ly(
    data  = plot_df,
    x     = ~bin_mids,
    y     = ~bin_counts,
    type  = "bar",
    color = ~type, 
    colors = plotly_colors_vector,
    marker = list(line = list(color = "#2e2e2e", width = 1),
                  width = ~bin_widths),
    text = ~hover_text,
    hoverinfo = "text",
    textposition = "none"
  ) %>%
    layout(
      title = paste(site, roi_name),
      xaxis = list(title = roi_name, range = c(min(bin_breaks), max(bin_breaks))),
      yaxis = list(title = "Frequency", type = y_axis_type_param),
      bargap = 0,
      showlegend = TRUE,
      shapes= sdev_shapes,
      annotations = sdev_annotations
    )
}

```


<script>
  document.getElementById('scaleInput').onchange = async function () {changePlotAxis()}
  
  async function changePlotAxis() {
    const scaleInputValue = document.getElementById('scaleInput').value;
    const plots = document.getElementsByClassName("js-plotly-plot");
    await Promise.all(
      Array.from(plots).map(el =>
        Plotly.relayout(el, { 'yaxis.type': scaleInputValue })
      )
    );
  }
  
</script>

<script>
  document.getElementById('sdLines').onchange = async function () { toggleLines() };
  
  async function toggleLines() {
    const isChecked = document.getElementById('sdLines').checked
    const plots = document.getElementsByClassName("js-plotly-plot");
    
    await Promise.all(
        Array.from(plots).map(el => { 
          const currentLayout = el.layout;

          const updatedShapes = (currentLayout.shapes || []).map(shape => ({
              ...shape,
              visible: isChecked
          }));

          const updatedAnnotations = (currentLayout.annotations || []).map(annotation => ({
              ...annotation,
              visible: isChecked
          }));

          return Plotly.relayout(el, {
              shapes: updatedShapes,
              annotations: updatedAnnotations
          });
        })
      );
  }
  
</script>

<script>
  document.getElementById('legend').onchange = async function () { toggleLegend() };
  
  async function toggleLegend() {
    const isChecked = document.getElementById('legend').checked
    const plots = document.getElementsByClassName("js-plotly-plot");
    
    await Promise.all(
        Array.from(plots).map(el => { 
          const currentLayout = el.layout;

          return Plotly.relayout(el, {
             showlegend: isChecked
          });
        })
      );
  }
  
</script>
```{r generate_plots, results='asis'}
# Plotly plots from within a loop don't render, using the following fix:
# https://stackoverflow.com/questions/49990653/plotly-plot-doesnt-render-within-for-loop-of-rmarkdown-document

output <- list()
sd_multiples <- c(2, 5)
y_axis_type_param <- 'linear'

if (length(rois_list) == 0) {
  cat("### No valid ROIs found in the provided list.\n")
} else {
  for (roi_name in rois_list) {
    # ROI Header
    txt <- paste0("## ROI: ", roi_name, "\n\n")
    output[[length(output) + 1L]] <- txt
    
    if (!roi_name %in% names(data_table)) {
      cat(paste0("### Skipping ROI: ", roi_name, " *(Column not found in data)*\n\n"))
      next
    }
    
    DATA <- data_table[[roi_name]]
    if (all(is.na(DATA))) {
      cat(paste0("### Skipping ROI: ", roi_name, " *(All values missing after cleaning)*\n\n"))
      next
    }
    
    # Subject IDs (if available)
    SUBJECT_DATA <- if ("subjectID" %in% names(data_table)) data_table$subjectID else NULL
    
    # --- Compute Stats ---
    stats <- compute_roi_stats(DATA)
    
    # --- Prepare Data for Plot ---
    plot_df <- prepare_histogram_data(DATA, hbins, stats$mean, stats$sd, sd_multiples)
    plot_df <- add_hover_text(plot_df, DATA, SUBJECT_DATA, stats$mean, stats$sd, sd_multiples)
    
    # --- Generate Interactive Plot ---
    p <- generate_roi_plot(
      plot_df, 
      roi_name, 
      site, 
      unique(c(plot_df$bin_start, plot_df$bin_end)),
      stats$mean,
      stats$sd,
      y_axis_type_param,
      sd_multiples
      )
    
    # Add to output list
    output[[length(output) + 1L]] <- p
  }
}

# Render all plots sequentially
for (x in output) {
  if (inherits(x, "character")) {
    cat(x)
  } else {
    cat(htmltools::renderTags(as.widget(x))$html)
  }
}
```


```{r dependencies, echo=FALSE, message=FALSE, warning=FALSE}
# Attach the dependencies so that Plotly widgets render correctly when using renderTags()
deps <- lapply(
  Filter(f = function(x) inherits(x, "htmlwidget"), x = output),
  function(hw) htmltools::renderTags(hw)$dependencies
)

htmltools::attachDependencies(
  x = htmltools::tagList(),
  value = unlist(deps, recursive = FALSE)
)
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
